ptr-tidy: Automatic Rejuvenation of Raw Pointers in C++

### 2021-04-13 paper.tex
Go over finished draft

    \begin{abstract}
    Systems programming involves dynamically requesting memory that in the C++ language has to be manually managed by the programmer. 
    However, this leaves the opportunity for exploitable memory safety errors.
    Modern C++ recommends that smart pointer are used to automatically manage resources such as dynamic memory.
    However, many legacy programs may have been written before these features were introduces to the language.
    We developed the ptr-tidy tool in order to automatically rejuvenate such programs to use smart pointers.
    The tool uses the Clang and LLVM libraries in order to create a componentised and powerful design which could be reused by other refactoring or rejuvenation tools. 
    We present a case study and and apply the tool on a number of open-source C++ libraries to show that the tool can successfully rejuvenate non-trivial libraries and that the tool can be used on libraries from many areas of computer science. 
    We discuss in which circumstances the tool performs well and the current limitations.
    \end{abstract}

@@

    Most ordinary computer users demand that the machine they are using provides them with a responsive, secure and productive environment to complete their tasks.
    The programs that are most responsible for this are complex systems programs such as the underlying operating system, device drivers and web browsers.
    Systems programming involves memory management, that is dynamically requesting memory from the operating system to be used and managed by the program.
    This is done when the amount of memory that we need at runtime cannot be know at the time of compilation of the program.
    However, doing so creates the opportunity for memory safety errors~\cite{Dhurjati2003} which are notoriously challenging to avoid.
    A memory safety error can be one of:
    \begin{itemize}
        \item \textbf{Dangling pointers}, prematurely freeing an object that is still in use.
        \item \textbf{Double free}, deleting an object multiple times.
        \item \textbf{Never free}, never deleting an object.
    \end{itemize}
    There are other categories of memory safety errors such as buffer overflows which we do not consider in this paper.
    If there are memory safety errors in a program, there is the possibility that it could be exploited by a malicious party to crash the program, expose sensitive data or gain remote code execution capabilities~\cite{Afek2007}.

    Historically systems programs are mostly written in C, C++ or a combination of the two and these remain the dominant languages.
    They are lower level languages that involve manual memory management by the programmer.
    It is even more challenging to ensure that C and C++ programs contain no memory safety errors as evidenced by empirical evidence in industry, where in a presentation at the Linux Security Summit it is shown that in several popular projects such as Firefox, macOS, Ubuntu and Android all had over half of their CVEs attributed to issues with memory safety~\cite{Gaynor2020}.
    Similarly, in a presentation by Matt Miller, a security engineer at Microsoft, it is shown that around 70\% of their vulnerabilities that are addressed through a security updates are due to memory safety issues~\cite{Miller2019}. 
    This is at industry-leading companies who are renowned for hiring top talent and top developer education programs.

@@

    Unfortunately, simply not using these languages is not an option since their unmanaged nature makes them highly performant, and therefore the best option for systems such as a web browser.
    Engineers at both Microsoft~\cite{Thomas2019} and Mozilla~\cite{Hostfelt2019} converge on Rust~\cite{Balasubramanian2017} as a possible solution.
    Rust is a systems language that offers similar performance as C and C++~\cite{Lin2016}, however its linear type system and memory ownership model also guarantee memory safety.
    The rewriting of a program in a new language, namely from C++ to Rust, is a colossal undertaking, especially given that Rust has a reputation for being difficult to learn~\cite{Astrauskas2019}.
    Instead, modern C++ guidelines encourage the use of \emph{smart pointers}~\cite{Dimov2003} and the concept of \emph{RAII (Resource Acquisition Is Initialisation)}~\cite{Stroustrup2000} to create memory or resource handles which automatically eliminate entire classes of memory safety errors.
    Yet, we encounter further problems as C++ programs either do not actively follow this advice, or are \emph{legacy} programs that were written before these features were introduced to the language.

    We therefore propose a solution in the form of a tool, \emph{ptr-tidy}~\cite{Usov2021}, to analyse and automatically refactor programs to follow the recommended use of smart pointers where the analysis can determine that such a change is syntactically correct and does not change the runtime behaviour of the program.

    \begin{listing}
        \inputminted{c++}{code/leak.cpp}
        \inputminted{c++}{code/fix-leak.cpp}
        \caption{Example of a memory leak using manual management and how smart pointers correct this. Lines 4 and 5 show how interruptions in the program control flow can mistakenly cause memory leaks and how smart pointer correct this by automatically inserting all needed delete statements at compile time. All necessary headers are omitted for brevity}
        \label{lst:motivating-example}
    \end{listing}

    We provide a motivating example in \autoref{lst:motivating-example} where the first version of the function could have several control flow executions where the \texttt{Circle} object will leak.
    The example seems quite trivial, but can suddenly be quite non-trivial if the conditions in the \texttt{if} statement become calls into external libraries, which may or may not throw an exception.

@@

    Our work closely relates to work done around the field of \emph{source code rejuvenation}, as defined by \citet{Pirkelbauer2010}. 
    We exactly aim replace outdated coding patterns with newer, higher-level abstractions, as well as fitting the notion that this tool would be applied once to a codebase, rather than be a reoccurring task.
    We will however use the terms rejuvenation and refactoring interchangeably in this paper as it is a term that most developers are familiar with.

    We believe research in this area is going to increase, as languages move and evolve much quicker than can be supported in enterprise environments, resulting in accumulating technical debt. 
    This is supported by the fact that other relevant work has been completed on this subject.

    First and foremost, we have evidence that this is a problem that exists in industry through a paper by \citet{Wright2013} which describes a tool used at Google for running such rejuvenations across a large codebase. 
    \citet{Wright2013} show an example of running a simple rejuvenation of upgrading an API call.
    They use Clang for the parsing of code into an AST and traverse the tree to identify and perform simple refactorings.
    Their flexible design allows them to implement many different simple refactorings using the same tool, rather than the focus we have for a single specific refactoring goal.

    Secondly, there has been separate academic research into rejuvenation tools such as the work by \citet{Huck2016} and \citet{Kumar2012}.
    The former also uses the Clang and AST driven approach like \citet{Wright2013}, while the latter uses a representation called IPR~\cite{Dos2011}, which is a general and efficient data structure for representing C++ programs.
    We have not evaluated IPR as a representation compared to the Clang AST, but we can however comment that whilst we appreciate the aim of the project to create a efficient and compiler-independent representation, the greater amount of documentation that Clang has, as well as it being the choice of several other successful projects leads us to believe that it may be the more suitable and mature option for any such research.

    There has also been industry interest in automatic rejuvenation tools between languages rather than within the same language. 
    This is done either to increase the use of a new, improved language by creating creating tools to aid transition, or in order to reuse existing compilers.
    For example, the Kotlin language team created the J2K~\footnote{\url{https://github.com/JetBrains/kotlin/tree/master/j2k}} transpiler in order to entice Java developers to switch to a language with improved language features such as null safety~\cite{Mateus2019}. 
    Similarly, several new languages such as Typescript and Dart have been recently developed which were designed not to be used by themselves, but to be transpiled to the popular Javascript language.
    These languages aim to solve shortcomings of the output language, whilst reusing the existing mature and efficient compilers or interpreters.

@@

    Generally pointers are used to give the program access to a resource that cannot be directly included in the program itself, such as allocated memory or a file.
    Pointers can also be used to pass large resources to functions without needing to create a copy, which would have a performance impact.

    Pointers can be created for objects both on \emph{the stack and the heap}~\cite{Stroustrup2000}. 
    The stack is an small memory store automatically managed for us by the compiler and is suitable for placing small, fixed sized parameters and arguments, such as perhaps loop counter variables.
    However, when objects must live beyond the lifetime of the current function call or if we do not know the size of an object until runtime, then we must storage these objects on the heap, occasionally called the \emph{free store}. 
    Unlike the stack, the memory on the heap must be explicitly managed by the programmer.

    However Stroustrup claims that pointers to objects allocated on the free store are dangerous and a \emph{plain old pointer}, or raw pointer as we will refer to them, should not be used to represent ownership~\cite{Stroustrup2018}.

    Instead we can use \emph{smart pointers}~\cite{Dimov2003} and the concept of \emph{RAII (Resource Acquisition Is Initialisation)}~\cite{Stroustrup2000} to create resource handles which automatically eliminate resource leaks with no, or very little added overhead.
    RAII means that the management of a resource, namely its acquisition and release, is bound to its lifetime. 
    In the simplest case, this means that when a resource exists the scope of a function or method, it is released automatically.

    The first type of smart pointer are \emph{unique pointers} which have only a single unique owner at any time and therefore are ideal for resources which may not be copied, such as perhaps a global session dictionary.
    We can see how unique pointers and RAII work from out motivating example in  \autoref{lst:motivating-example}.
    Even though the \texttt{Circle} object is placed on the heap, the compiler can manage the smart pointer for us so that in all cases where the lifetime of the object ends, the object will be freed and ensures the resource will not leak. 
    The ownership of a unique pointer can be transferred between variables using \texttt{std::move} which moves the ownership from one variable to another, making the original reference invalid.
    % TODO insert example diagram possibly

@@

    As seen in \autoref{sec:background}, much similar research uses Clang and LLVM.
    LLVM~\cite{Lattner2004} is a compiler infrastructure project consisting of several compiler and toolchain technologies. 
    LLVM is centrally designed around a language independent intermediate representation (IR), which can be created by a number of language front-ends and is used by a number of architecture back-ends to generate machine code. 
    Program optimisations can then be applied by portable and reusable transformations on the IR.
    Any improvement in back-end code generation, as well as any new or improved IR optimisers can then benefit any of the supported language front-ends.
    It is therefore visible that this allows for a very modular and extensible design which can be seen in \autoref{fig:llvm-design}.

    Clang~\cite{Lattner2008} is such an LLVM front-end for C, C++ and a variety of other similar languages such as OpenGL.
    By using the LLVM infrastructure, it allows for faster and more efficient compilation of C++ compared to existing, older compilers such as the GNU Compiler Collection~\footnote{\url{https://gcc.gnu.org/}} (GCC)~\cite{Lattner2008}.
    However, most importantly for our research, the design of GCC makes it unsuitable for integration with other projects, such as analysis tools, whereas Clang allows access to several analysis tools as well as the Clang abstract syntax tree (AST) and parser internals via a standard API such as consumers and visitors~\cite{Duffy2014}.

    \begin{figure}
        \centering
        \inputminted{c++}{code/code-to-ast.cpp}
        \includegraphics{images/code-to-ast.pdf}
        \caption{Example of converting C++ code to Clang AST. All diagrams show equivalent representation of a small C++ code snippet.}
        \label{fig:code-to-ast-example}
    \end{figure}

    Further more, Clang also explicitly provides the \emph{LibTooling} C++ library for the creation of tools that leverage the Clang parsing front-end and AST.
    The Clang AST is different from ASTs produced by some other compilers in that it closely resembles both the written C++ code and the C++ standard~\footnote{\url{https://clang.llvm.org/docs/IntroductionToTheClangAST.html}}.
    For example, we can see in \autoref{fig:code-to-ast-example} that the AST stores much syntactic as well as semantic information in the form of source code locations and variable initialisation styles.
    The LLVM authors remark that this makes it very suitable for refactoring tools.

    The LLVM IR represents a language-independent, static single assignment (SSA)~\cite{Rosen1988} representation of the program.
    Variables in in SSA form can only be assigned once, which allows us to efficiently generate compact definition-usage graphs, which enable developers to simplify many dataflow optimizations and enables fast, flow-insensitive algorithms~\cite{Lattner2004}.

@@

    The first capability our tool would need is to be able to parse C++ code.
    Doing this ourselves would be an undertaking that would take a significant amount of time in order to create a complete parser due to the size and complexity of the C++ language, therefore it was decided early to use an existing parser.
    The information in \autoref{subsec:clang-and-llvm} outlines why Clang is the most suitable parser to use, with the other benefit that it brings such as \emph{LibTooling}.

@@

    Next, our tool would need to be able to identify specifically which pointers are suitable for refactoring.
    An \emph{unique pointer} is a smart pointer object that represent a single unique owner of a region of memory at any time. 
    We will therefore need to develop a static analysis algorithm to identify the maximum number of owners a region of memory has throughout all points of a program. 
    If a memory region has at most a single unique owner at a time, we can refactor the initialising raw pointer to a unique pointer, otherwise the pointer can be refactored to a shared pointer.    
    This analysis will need to ensure that its result is correct, and will inevitably have to make some conservative assumptions about the program.

@@

    A key intuition is that a region of memory, subsequent to its initialisation, is only able to gain additional owners if access to the region of memory is \emph{shared} to another variable, that is if a pointer to this memory is copied to a another pointer, at which point the original owner is longer the unique owner.
    If we copy the pointer to a global variable, the variable could suddenly be read by any other running threads in the program, so we have to take a conservative decision that the memory no longer has a single unique owner.
    If we copy the pointer to a local variable, then if the original variable remains in use, the memory will have two unique owners. 
    However if the original pointer is not used beyond being copied, this could actually be modelled as a \emph{transfer} of ownership, and the memory will still have a single unique owner at all points of the program. 
    We show this intuition in \autoref{fig:memory-pointers}, where the memory region \texttt{0001} is not suitable for being managed by a unique pointer, whilst region \texttt{0003} is.

    We can formalise our key intuition by introducing the concept of \emph{escape analysis}.
    Escape analysis is used to determine whether an object \emph{escapes} or is accessible from outside the method or thread that created the object~\cite{Choi1999}.
    It has been used in languages such as Java~\cite{Choi1999} to determine whether an object which is created within a method escapes.
    If it does not escape, meaning it is only used locally in the method, then it can be allocated on the stack rather than the heap as a performance optimisation.
    LLVM also uses escape analysis for a similar kind of optimisation~\footnote{\url{https://github.com/llvm/llvm-project/blob/main/llvm/lib/Analysis/CaptureTracking.cpp}}.

    It is now visible that the invariant of the unique pointer are only violated when:
    \begin{itemize}
        \item a pointer escapes, as then we do not know where or how the pointer may be used, so we must take a conservative approach and assume the memory region no longer has at most a single unique owner.
        \item a pointer is does not escape but access to the memory is shared rather than transferred to a local variable. 
    \end{itemize}

@@

    The existing escape analysis in LLVM is implemented to operate on LLVM IR code, as the IR representation, as mentioned in \autoref{subsec:clang-and-llvm}, is well suited for efficient analysis algorithms. 
    However, much of the existing work from \autoref{sec:related-work} which use Clang perform both their analyses and their source code refactoring on the Clang AST.
    We however propose an improved novel design which uses the Clang AST for parsing and refactoring and the LLVM IR for simple and efficient analysis.

    For the same input file, we can use the \emph{LibTooling} library to generate an AST and IR in-memory representation.
    Then, for any variable declaration (\emph{VarDecl}) in the Clang AST which allocates new memory on the heap, we locate this declaration in the IR.
    This is possible as we can instruct Clang to not discard variable names using a \texttt{fno-discard-value-names} flags, and as the combination of function signature and variable name in the translation unit (\emph{cpp file}) must be unique according to the One-definition rule~\cite{C++17}.

@@

    In \autoref{fig:code-to-ast-example}, given the variable declaration for the variable \texttt{a}, then as it is not a global variable, we can walk its parent nodes until we reach a \emph{FunctionDecl}.
    We are now able to use the function name to find the corresponding function declaration in the IR module, which can be seen in  \autoref{lst:code-to-ir}.
    Note that although the \texttt{main} function name corresponds exactly, other function names in the IR module will have mangled names~\footnote{\url{https://www.ibm.com/support/knowledgecenter/en/ssw_ibm_i_72/rzarg/name_mangling.html}}.
    We again use Clang library functions to mangle function names if necessary so they can be found in the IR. 
    Given a function in the IR, we can use its symbol table to use the declaration name to find the IR value on which we will perform our analysis.

@@

    The analysis gives a result of whether the IR value escapes, a boolean \texttt{true} or \texttt{false}.
    For a \texttt{false} result, the value declaration is refactored into a unique pointer, otherwise a shared pointer.
    However, the rewriting is identical for both cases with just the string tokens that we replace for the original type differing.
    The rewriting was developed using a rule based approach, using test driven development (TDD) as this allowed rapid but correct development by continuously being able to check that additional created rules still cause all the tests to pass, causing no regressions. 
    The five cases of code (\emph{or rewriting rules}) that needs to be rewritten are:
    \begin{itemize}
        \item Declaration types.
        \item Variable initialisation and assignment expressions.
        \item Pointer delete statements.
        \item Function return types.
        \item Pointer Usages.
    \end{itemize}

    Rewriting declaration types are the easiest of all. We can directly replace the type of the variable declaration node that we already have access to. 
    For example for \mintinline{c++}{int *ptr}, we can extract the type of the variable, and format it into the string \mintinline{c++}{"std::shared_ptr<{}>"} or \mintinline{c++}{"std::unique_ptr<{}>"}, where the \texttt{\{\}} represent the replacement field.

    Initialisation and assignment expressions are less simple due to two reason.
    First of all, the initialisation of a variable does not necessarily have to occur immediately at the point of declaration.
    Therefore we need to find all the points of initialisation or value assignment to a variable in the AST.
    Secondly, there are many different ways to initialise a variable~\cite[p.~196]{C++17}, with the most common initialisation methods using:
    \begin{itemize}
        \item Braced-lists \mintinline{c++}{std::string s{}}.
        \item Equal sign \mintinline{c++}{std::string s = "hello"}
        \item parentheses expression list \mintinline{c++}{std::string s("hello")} 
    \end{itemize}
    For pointer variables, this is convoluted even more depending on if the pointer gets initialised using an existing pointer \mintinline{c++}{int* a = nullptr; int *b = a;}, or if it gets initialised using a new statement \mintinline{c++}{auto p = new Circle{Point{0,0},10}}.
    In the case of a new statement, we can initialise the rewritten smart pointer using a value initialiser \mintinline{c++}{std::make_unique<>()} or the shared pointer equivalent. 
    Otherwise, we can create a smart pointer from a pointer value using by the default equal sign copy initialiser or parentheses expression list.

>>>>

    \begin{abstract}
    Systems programming involves dynamically requesting memory which in the C++ language has to be manually managed by the programmer. 
    However, this leaves the opportunity for exploitable memory safety errors.
    Modern C++ recommends that smart pointers are used to automatically manage resources such as dynamic memory.
    However, many legacy programs may have been written before these features were introduced to the language.
    We develop the ptr-tidy tool in order to automatically rejuvenate such programs to use smart pointers.
    The tool uses the Clang and LLVM libraries in order to create a componentised and powerful design which could be used by other refactoring or rejuvenation tools. 
    We present a case study and also apply the tool on a number of open-source C++ libraries from a variety of areas in computer science to show that the tool can successfully rejuvenate non-trivial libraries.
    We discuss the circumstances in which the tool performs well and the current limitations.
    \end{abstract}

@@

    Most ordinary computer users demand that the machine they are using provides them with a responsive, secure and productive environment to complete their tasks.
    The programs that are most responsible for this are complex systems programs such as the underlying operating system, device drivers and web browsers.
    Systems programming involves memory management, that is dynamically requesting memory from the operating system to be used and managed by the program.
    This is done as the amount of memory that we need at runtime cannot be know at the time of compilation of the program.
    However, doing so creates the opportunity for memory safety errors~\cite{Dhurjati2003}.
    A memory safety error can be one of:
    \begin{itemize}
        \item \textbf{Dangling pointers}, prematurely freeing memory that is still in use.
        \item \textbf{Double free}, freeing memory multiple times.
        \item \textbf{Never free}, never freeing memory.
    \end{itemize}
    There are other categories of memory safety errors such as buffer overflows which we do not consider in this paper.
    If there are memory safety errors in a program, there is the possibility that it could be exploited by a malicious party to crash the program, expose sensitive data or gain remote code execution capabilities~\cite{Afek2007}.

    Historically, systems programs are mostly written in C, C++ or a combination of the two and these remain the dominant languages.
    They are lower level languages that involve manual memory management by the programmer.
    It is challenging to ensure that C and C++ programs contain no memory safety errors as evidenced by empirical data in industry, where in a presentation at the Linux Security Summit it was shown that in several popular projects such as Firefox, macOS, Ubuntu and Android all had over half of their CVEs~\footnote{\url{https://cve.mitre.org/}} attributed to issues with memory safety~\cite{Gaynor2020}.
    Similarly, in a presentation by Matt Miller, a security engineer at Microsoft, it is shown that around 70\% of their vulnerabilities that are addressed through security updates are due to memory safety issues~\cite{Miller2019}. 
    % This is at industry-leading companies who are renowned for hiring top talent and with top developer education programs.

@@

    Unfortunately, simply not using C or C++ is not an option since their unmanaged nature allows for high performance, and therefore the best option for systems with strict or high requirements.
    Engineers at both Microsoft~\cite{Thomas2019} and Mozilla~\cite{Hostfelt2019} converge on Rust~\cite{Balasubramanian2017} as a possible solution.
    Rust is a systems language that offers similar performance as C and C++~\cite{Lin2016}, however its linear type system and memory ownership model guarantee memory safety.
    The rewriting of a program in a new language, namely from C++ to Rust, is a colossal undertaking, especially given that Rust has a reputation for being difficult to learn~\cite{Astrauskas2019}.
    Instead, modern C++ guidelines encourage the use of \emph{smart pointers}~\cite{Dimov2003} and the concept of \emph{RAII (Resource Acquisition Is Initialisation)}~\cite{Stroustrup2000} to create compiler managed resource handles.
    Their correct use eliminates the three previously identified classes of memory safety errors.
    Yet, we encounter further problems as C++ programs either do not actively follow this advice, or are \emph{legacy} programs which were written before these features were introduced to the language.

    We therefore propose a solution in the form of a tool, \emph{ptr-tidy}~\cite{Usov2021}, to analyse and automatically refactor programs to follow the recommended use of smart pointers where the analysis can determine that such a change is correct and does not change the runtime behaviour of the program.

    \begin{figure}
        \centering
        \inputminted{c++}{code/leak.cpp}
        \includegraphics{images/arrow.pdf}
        \inputminted{c++}{code/fix-leak.cpp}
        \caption{Example of a memory leak using manual management and how smart pointers correct this problem. Lines 4 and 5 show how interruptions in the program control flow can mistakenly cause memory leaks as the \texttt{delete} statement is never reached. Smart pointers correct this by automatically inserting delete statements at all necessary locations at compile time. All necessary headers are omitted for brevity}
        \label{lst:motivating-example}
    \end{figure}

    We provide a motivating example in \autoref{lst:motivating-example} where the initial version of the function has several control flow statements that could cause the \texttt{Circle} object to never be freed.
    The example can seem trivial as all the possible executions of the program are easy to visualise, but can suddenly be rather non-trivial if the conditions in the \texttt{if} statement become calls into external libraries, which may throw an exception and cause the \texttt{delete} statement to again not be reached.

@@

    Our work closely relates to work done around the field of \emph{source code rejuvenation}, as defined by \citet{Pirkelbauer2010}. 
    We exactly aim to replace outdated coding patterns with newer, higher-level abstractions, as well as fitting the notion that this tool would be applied once to a codebase, rather than be a reoccurring task.
    We will however use the terms rejuvenating and refactoring interchangeably in this paper as it is a term that most developers are more familiar with.

    We believe research in this area is going to increase, as languages move and evolve much quicker than can be supported in enterprise environments~\cite{Overbey2009, Favre2005}, resulting in accumulating technical debt. 
    This is supported by the fact that other relevant work has been completed on this subject.

    First and foremost, we have evidence that this is a problem that exists in industry through a paper by \citet{Wright2013} which describes a tool used at Google for running such rejuvenations across a large C++ codebase. 
    \citet{Wright2013} show an example of running a simple rejuvenation of upgrading an API call.
    They use Clang for the parsing of code into a syntax tree and traverse the tree to identify and perform simple refactorings.
    Their flexible design allows them to implement many different simple refactorings using the same tool, rather than the focus we have for a single specific refactoring goal.

    Secondly, there has been separate academic research into rejuvenation tools such as the work by \citet{Huck2016} and \citet{Kumar2012}.
    The former also uses the Clang driven approach like \citet{Wright2013}, while the latter uses a representation called IPR~\cite{Dos2011}, which is a general and efficient data structure for representing C++ programs.
    We have not evaluated IPR as a representation compared to the use of Clang, but we can  comment that whilst we appreciate the aim of the project to create a efficient and compiler-independent representation, the greater amount of documentation for Clang, as well as being the choice of several other successful projects leads us to believe that it may be the more suitable and mature option for any such research.

    There has also been industry interest in automatic rejuvenation tools \emph{between} languages rather than \emph{within} the same language. 
    This is done either to increase the use of a new, improved language by creating tools to aid transition, or in order to reuse existing compilers.
    For example, the Kotlin language team created the J2K~\footnote{\url{https://github.com/JetBrains/kotlin/tree/master/j2k}} transpiler in order to entice Java developers to switch to a language with improved language features such as null safety~\cite{Mateus2019}. 
    Similarly, several new languages such as Typescript~\footnote{\url{https://www.typescriptlang.org/}} and Dart~\footnote{\url{https://dart.dev/}} have been recently developed which were designed not to be used by themselves, but to be transpiled to the popular Javascript language.
    These languages aim to solve shortcomings of the output language, whilst reusing the existing mature and efficient compilers or interpreters.

@@

    Generally, pointers are used to give a program access to a resource that cannot be directly included in the program itself, such as allocated memory or a file.
    Pointers can also be used to pass large resources to functions without needing to create a copy, which would have a performance impact.

    Pointers can be created for objects both on \emph{the stack and the heap}~\cite{Stroustrup2000}. 
    The stack is a small memory store automatically managed for us by the compiler and is suitable for placing object whose size is known at compile time, such as simple integers like loop counter variables.
    However, when objects must live beyond the lifetime of the current function call or if we do not know the size of an object until runtime, then we must store these objects on the heap, also called the \emph{free store}. 
    Unlike the stack, the memory on the heap must be explicitly managed by the programmer.

    However Stroustrup claims that pointers to objects allocated on the free store are dangerous and a plain old pointer, or \emph{raw pointer} as we will refer to them, should not be used to represent ownership of such objects.~\cite{Stroustrup2018}.

    Instead we can use \emph{smart pointers}~\cite{Dimov2003} and the concept of \emph{RAII (Resource Acquisition Is Initialisation)}~\cite{Stroustrup2000} to create resource handles with no, or very little added overhead.
    RAII means that the management of a resource, namely its acquisition and release, is bound to its lifetime. 
    In the simplest case, this means that when a resource exits the scope of a function or method, it is released automatically.
    We can see how smart pointers and RAII work in our motivating example in  \autoref{lst:motivating-example}.
    Even though the \texttt{Circle} object is placed on the heap, the compiler can manage the smart pointer for us so that in all the cases where the lifetime of the object ends, the object will be released.

    The first type of smart pointer are \emph{unique pointers}, which can have only a single unique owner at any time and therefore are ideal for resources which may not be copied.
    The resource managed by a unique pointer is released when the current owner exits the scope of the function or method it is contained in.
    The ownership of a unique pointer can be transferred between variables using \texttt{std::move} which \emph{moves} the ownership from one variable to another, making the original owner invalid.
    % TODO insert example diagram possibly

@@

    As seen from \autoref{sec:related-work}, much similar research uses Clang and thus LLVM.
    LLVM~\cite{Lattner2004} is a compiler infrastructure project consisting of several compiler and toolchain technologies. 
    LLVM is centrally designed around a language independent intermediate representation (IR), which can be created by a number of language front-ends and is used by a number of back-ends to generate machine code. 
    Programs can then be optimised by applying portable and reusable transformations on the IR.
    Any improvement in back-end machine code generation, as well as any new or improved IR optimisations therefore benefit all of the language front-ends.
    It is therefore visible that this allows for a very modular and powerful design which can be seen in \autoref{fig:llvm-design}.

    Clang~\cite{Lattner2008} is a LLVM language front-end for C, C++ and a variety of other extensions such as OpenGL.
    By using the LLVM infrastructure, it allows for faster and more efficient compilation of C++ compared to existing, older compilers such as the GNU Compiler Collection~\footnote{\url{https://gcc.gnu.org/}} (GCC)~\cite{Lattner2008}.
    However, most importantly for our research, the design of GCC makes it unsuitable for integration in other projects such as analysis tools, whereas Clang allows access to several analysis tools as well as the Clang abstract syntax tree (AST) and parser internals via a standard API such as consumers and visitors~\cite{Duffy2014}.

    \begin{figure}
        \centering
        \inputminted{c++}{code/code-to-ast.cpp}
        \includegraphics{images/code-to-ast.pdf}
        \caption{Example of converting C++ code to the Clang AST. The Clang AST is rich in information and could be used to reconstruct the original C++ code snippet.}
        \label{fig:code-to-ast-example}
    \end{figure}

    Further more, Clang also provides the \emph{LibTooling} C++ library for the creation of tools that leverage the Clang parsing front-end and AST.
    The Clang AST is different from ASTs produced by some other compilers in that it closely resembles the written C++ code~\footnote{\url{https://clang.llvm.org/docs/IntroductionToTheClangAST.html}}.
    For example, we can see in \autoref{fig:code-to-ast-example} that the AST stores much syntactic as well as semantic information in the form of source code locations and variable initialisation styles.
    The LLVM authors remark that this makes it very suitable for refactoring tools.

    The LLVM IR is a language-independent, static single assignment (SSA)~\cite{Rosen1988} representation of a program.
    Variables in SSA form can only be assigned once, which allows us to efficiently generate compact definition-usage graphs, which enable developers to simplify many dataflow optimizations and enables fast, flow-insensitive algorithms~\cite{Lattner2004}.

    \vskip50pt

@@

    The first capability our tool needs is to be able to parse C++ code.
    Doing this ourselves would be an undertaking that would take a significant amount of time in order to create a complete parser due to the size and complexity of the C++ language. Therefore it was decided early to use an existing parser.
    The information in \autoref{subsec:clang-and-llvm} outlines why Clang is the most suitable parser to use, with the other benefits that it brings such as the \emph{LibTooling} library.

@@

    Next, our tool would need to be able to identify specifically which pointers are suitable for refactoring into smart pointers.
    An \emph{unique pointer} is a smart pointer that represents a single unique owner of a region of memory at any point in time.
    We will therefore need to develop a static analysis algorithm to identify the maximum number of owners a region of memory has throughout all points of the program. 
    If a memory region has at most a single unique owner, then we can refactor the initialising raw pointer to be a unique pointer, and otherwise to a shared pointer.    
    This analysis will need to ensure that its result is correct, and will inevitably have to make some conservative assumptions about the program.

@@

    A key intuition is that a region of memory, subsequent to its initialisation, is only able to gain additional owners if access to the region of memory is \emph{shared} to another variable, that is if a pointer to this memory is copied to a another pointer, at which point the original owner is no longer the unique owner.
    If we copy the pointer to a global variable, the variable could suddenly be read by any other running threads in the program, so we have to take a conservative decision that the memory no longer has a single unique owner.
    If we copy the pointer to a local variable, then if the original variable remains in use, the memory will have two unique owners. 
    However if the original pointer is not used beyond being copied, this could actually be modelled as a \emph{transfer} of ownership, and the memory will still have a single unique owner at all points of the program. 
    We show this intuition in \autoref{fig:memory-pointers}, where the memory region \texttt{0001} is not suitable for being managed by a unique pointer, whilst region \texttt{0003} is.

    We can formalise our key intuition by introducing the concept of \emph{escape analysis}.
    Escape analysis is used to determine whether an object \emph{escapes} or is accessible from outside the method or thread that created the object~\cite{Choi1999}.
    It has been used in languages such as Java~\cite{Choi1999} to determine whether an object which is created within a method escapes.
    If it does not escape, meaning it is only used locally in the method, then the object can be allocated on the stack rather than the heap as a performance optimisation.
    LLVM also uses escape analysis for a similar kind of optimisation~\footnote{\url{https://github.com/llvm/llvm-project/blob/main/llvm/lib/Analysis/CaptureTracking.cpp}}.

    It is now visible that the invariant of the unique pointer are only violated when:
    \begin{itemize}
        \item a pointer escapes, as then we do not know where or how the pointer may be used, so we must take a conservative approach and assume the memory region no longer has at most a single unique owner.
        \item a pointer does not escape but access to the memory is shared rather than transferred between local variables. 
    \end{itemize}

@@

    The existing escape analysis in LLVM is implemented to operate on LLVM IR code, as the IR representation, as mentioned in \autoref{subsec:clang-and-llvm}, is well suited for efficient analysis algorithms. 
    However, much of the existing work in \autoref{sec:related-work} use Clang and the Clang AST to perform their parsing, analyses and their source code refactoring..
    We however propose an improved novel design which uses the Clang AST for parsing and refactoring and the LLVM IR for efficient analysis.

    For the same input file, we can use the \emph{LibTooling} library to generate an AST and IR in-memory representation.
    Then, for any variable declaration (\texttt{VarDecl}) in the Clang AST which allocates new memory on the heap, we locate this declaration in the IR.
    This is possible as we can instruct Clang to not discard variable names using a \texttt{fno-discard-value-names} flag, and as the combination of function signature and variable name in the translation unit must be unique according to the One-definition rule~\cite{C++17}.

@@

    In \autoref{fig:code-to-ast-example}, given the variable declaration for the variable \texttt{a}, then as it is not a global variable, we can visit its parent nodes in the AST until we reach a \texttt{FunctionDecl}.
    We can then use the name of the \texttt{FunctionDecl} to find the corresponding function declaration in the IR module, which can be seen in  \autoref{lst:code-to-ir}.
    Note that although the \texttt{main} function name corresponds exactly, other function names in the IR module will have mangled names~\footnote{\url{https://www.ibm.com/support/knowledgecenter/en/ssw_ibm_i_72/rzarg/name_mangling.html}}.
    We again use Clang library functions to mangle function names if necessary so they can be found in the IR. 
    Given the IR function declaration, we can find the corresponding IR value of our initial variable \texttt{a} by finding the name of the variable inside its symbol table.
    This IR value can then be used to perform our analysis.

@@

    The analysis gives a result of whether the IR value escapes, a boolean \texttt{true} or \texttt{false}.
    For a \texttt{false} result, meaning the value does not escape, the variable declaration is refactored into a unique pointer, and otherwise a shared pointer.
    However, the rewriting of the source code is identical for both cases with just the string tokens that replace the original source code differing.
    The rewriting was developed using a rule based approach, using test driven development (TDD) as this allowed rapid but correct development by continuously being able to check that additional created rules still cause all the tests to pass, causing no regressions. 
    The five cases of code (\emph{or the rewriting rules}) that needs to be rewritten are:
    \begin{itemize}
        \item The type of a variable declaration
        \item Variable initialisation or assignment expressions.
        \item Pointer delete statements.
        \item Function return types.
        \item Function call expressions that expect raw pointers.
    \end{itemize}

    Rewriting the type of a variable declaration is the easiest of all. We can directly replace the type of the variable declaration node that we already have access to. 
    For example for \mintinline{c++}{int *ptr}, we can extract the type of the variable, and format it into the string \mintinline{c++}{"std::shared_ptr<{}>"} or \mintinline{c++}{"std::unique_ptr<{}>"}, where \texttt{\{\}} represent the replacement field, which in this case will be formatted with the string \mintinline{c++}{int}.

    Initialisation and assignment expressions are less simple due to two reasons.
    First of all, the initialisation of a variable does not necessarily have to occur immediately at the point of declaration.
    Therefore we need to find all the points of initialisation or value assignment to a variable in the AST.
    Secondly, there are many different ways to initialise a variable~\cite[p.~196]{C++17}, with the most common initialisation methods using:
    \begin{itemize}
        \item Braced-lists \mintinline{c++}{std::string s{}}.
        \item Equal sign \mintinline{c++}{std::string s = "hello"}
        \item parentheses expression list \mintinline{c++}{std::string s("hello")} 
    \end{itemize}
    For pointer variables, this is convoluted even more depending on if the pointer gets initialised using an existing pointer \mintinline{c++}{int* a = nullptr; int *b = a;}, or if it gets initialised using a \mintinline{c++}{new} statement \mintinline{c++}{auto p = new Circle{Point{0,0},10}}.
    In the case of a new statement, we can initialise the rewritten smart pointer using a value initialiser \mintinline{c++}{std::make_unique<>()} or the shared pointer equivalent. 
    Otherwise, we can initialise the rewritten smart pointer from an existing pointer by using the default equal sign copy initialiser or parentheses expression list.
